Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-01 paper/paper.tex
DFA parser wip

@@

\newcommand{\simulsubst}[2]{#1\{#2\}}
\newcommand{\subst}[3]{\simulsubst {#1} {#2/#3}}
\newcommand{\el}{\mathsf{el}}
\newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
\newcommand{\lamb}[2]{\lambda #1.\, #2}
\newcommand{\lamblto}[2]{\lambda^{{\lto}} #1.\, #2}
\newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
\newcommand{\dlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
\newcommand{\withlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
\newcommand{\app}[2]{#1 \, #2}
\newcommand{\applto}[2]{#1 \mathop{{}^{\lto}} #2}
\newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
\newcommand{\PiTy}[3]{\textstyle\prod (#1 : #2). #3}
\newcommand{\SigTy}[3]{\textstyle\sum (#1 : #2). #3}
\newcommand{\LinPiTy}[3]{\textstyle\bigamp (#1 : #2). #3}
\newcommand{\LinSigTy}[3]{\textstyle\bigoplus (#1 : #2). #3}
%% \newcommand{\DepWith}[2]{{\textstyle\bigamp}\limits_{#1}{#2}
%% \newcommand{\DepPlus}[2]{{\textstyle\bigoplus}\limits_{#1}{#2}
\newcommand{\GrTy}{\mathsf{Gr}}

@@

    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
    %
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A e x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \and
    %
    \inferrule
    {\Gamma \vdash M : \ltonl A}
    {\Gamma ; \cdot \vdash M : A}
    %
    \and
    %
    \inferrule
    { \Gamma ; \Delta \vdash e : A \\
      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {e}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
  \end{mathpar}
  \caption{Linear typing}
  \label{fig:linsyntax}
\end{figure}

@@

@@

\newcommand{\states}{\mathsf{states}}
\newcommand{\labelt}{\mathsf{label}}
\newcommand{\transitions}{\mathsf{transitions}}
\newcommand{\epstransitions}{\epsilon\mathsf{transitions}}
\newcommand{\isAcc}{\mathsf{isAcc}}
\newcommand{\init}{\mathsf{init}}
\newcommand{\src}{\mathsf{src}}
\newcommand{\dst}{\mathsf{dst}}
\newcommand{\epssrc}{\epsilon\mathsf{src}}
\newcommand{\epsdst}{\epsilon\mathsf{dst}}

@@

\begin{itemize}
\item $D.\states$ --- a finite set of states
\item $D.\init$ --- the initial state
\item $D.\isAcc$ --- a decidable proposition over $D.\states$ denoting which
      states are accepting
\item $D.\delta$ --- a transition function. Given a character $c$ and a state
        $s$, $D.\delta_{c}(s)$ is the state for which there is a transition $s \overset{c}{\to} D.\delta_{c}(s)$
\end{itemize}

The linear type of traces through $D$ is given in \cref{fig:dfatrace}. These
traces are either terminated in an accepting state with the $\nil$ constructor,
or the $\cons$ constructor builds a trace out of state $s$ by linearly combining
a parse of some character $c$ with a
trace out of the state $D.\delta_{c}(s)$.

\begin{figure}
\begin{align*}
\data &~\Trace_{D} : (s : D.\states) \to L~\where\\
      & \nil : \ltonl {\left( \LinPiTy {s} {D.\states} {\LinPiTy {\mathsf{acc}} {D.\isAcc(s)} {\Trace_{D}~s}} \right)}  \\
      & \cons : \ltonl {\left( \LinPiTy {c} {\Sigma} {\LinPiTy {s} {D.\states} {c \lto \Trace_{D}~(D.\delta_{c}(s)) \lto \Trace_{D}(s)}} \right)}
\end{align*}
\caption{Traces through the DFA $D$ as an indexed inductive type}
\label{fig:dfatrace}
\end{figure}

@@

\section{Semantics and Metatheory}
\label{sec:semantics-and-metatheory}

\subsection{Formal Grammars as Linear Types}
\label{sec:formaltype}
\max{this should be in its own section on semantics}

@@

Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
depth the linear semantics.

\begin{enumerate}
  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
        $\sem X : \sem \Gamma \to \Set_{i}$
  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$ \pedro{we have not defined what sections are}
  \item Linear contexts $\Gamma \vdash \Delta$ and types
        $\Gamma \vdash A : L_{i}$ both denote families of grammars
        $\sem \Gamma \to \Gr_{i}$
  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
        transformers
        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
\end{enumerate}

>>>>

\newcommand{\isTy}{\textrm{ type}}
\newcommand{\isCtx}{\textrm{ ctx}}
\newcommand{\isSmall}{\textrm{ small}}
\newcommand{\isLinTy}{\textrm{ lin. type}}
\newcommand{\isLinCtx}{\textrm{ lin. ctx.}}

@@

\newcommand{\simulsubst}[2]{#1\{#2\}}
\newcommand{\subst}[3]{\simulsubst {#1} {#2/#3}}
\newcommand{\el}{\mathsf{el}}
\newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
\newcommand{\lamb}[2]{\lambda #1.\, #2}
\newcommand{\lamblto}[2]{\lambda^{{\lto}} #1.\, #2}
\newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
\newcommand{\dlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
\newcommand{\withlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
\newcommand{\app}[2]{#1 \, #2}
\newcommand{\applto}[2]{#1 \mathop{{}^{\lto}} #2}
\newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
\newcommand{\PiTy}[3]{\textstyle\prod (#1 : #2). #3}
\newcommand{\SigTy}[3]{\textstyle\sum (#1 : #2). #3}
\newcommand{\LinPiTy}[3]{\textstyle\bigamp (#1 : #2). #3}
\newcommand{\LinPiTyLimit}[3]{\bigwith\limits_{#1 : #2} #3}
\newcommand{\LinSigTy}[3]{\textstyle\bigoplus (#1 : #2). #3}
\newcommand{\LinSigTyLimit}[3]{\bigoplus\limits_{#1 : #2} #3}
%% \newcommand{\DepWith}[2]{{\textstyle\bigamp}\limits_{#1}{#2}
%% \newcommand{\DepPlus}[2]{{\textstyle\bigoplus}\limits_{#1}{#2}
\newcommand{\GrTy}{\mathsf{Gr}}

@@

    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
    %
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A e x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \and
    %
    \inferrule
    {\Gamma \vdash M : \ltonl A}
    {\Gamma ; \cdot \vdash M : A}
    %
    \and
    %
    \inferrule
    { \Gamma ; \Delta \vdash e : A \\
      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {e}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
  \end{mathpar}
  \caption{Linear typing}
  \label{fig:linsyntax}
\end{figure}

@@

Through inclusion of this axiom, $! : \stringg \to \top$ is an isomorphism, and
thus also a monomorphism. That is, $\stringg$ is unambiguous.

@@

\newcommand{\states}{\mathsf{states}}
\newcommand{\labelt}{\mathsf{label}}
\newcommand{\transitions}{\mathsf{transitions}}
\newcommand{\epstransitions}{\epsilon\mathsf{transitions}}
\newcommand{\isAcc}{\mathsf{isAcc}}
\newcommand{\init}{\mathsf{init}}
\newcommand{\src}{\mathsf{src}}
\newcommand{\dst}{\mathsf{dst}}
\newcommand{\parse}{\mathsf{parse}}
\newcommand{\print}{\mathsf{print}}
\newcommand{\epssrc}{\epsilon\mathsf{src}}
\newcommand{\epsdst}{\epsilon\mathsf{dst}}

@@

\begin{itemize}
\item $D.\states$ --- a finite set of states
\item $D.\init$ --- the initial state
\item $D.\isAcc$ --- a decidable proposition over $D.\states$ denoting which
      states are accepting
\item $D.\delta$ --- a deterministic transition function. Given a character $c$ and a state
        $s$, $D.\delta_{c}(s)$ is the state for which there is a transition $s \overset{c}{\to} D.\delta_{c}(s)$
\end{itemize}

\begin{figure}
\begin{align*}
\data &~\Trace_{D} : (s : D.\states) (b : \Bool) \to L~\where\\
      & \nil : \ltonl {\left( \LinPiTyLimit {s} {D.\states} {\Trace_{D}(s , D.\isAcc(s))} \right)}  \\
      & \cons : \ltonl {\left( \LinPiTyLimit {c} {\Sigma} {\LinPiTyLimit {s} {D.\states} {c \lto \LinSigTyLimit {b}{\Bool}{\Trace_{D}~(D.\delta_{c}(s), b)} \lto \Trace_{D}(s, b)}} \right)}
\end{align*}
\caption{Traces through the DFA $D$ as an indexed inductive type}
\label{fig:dfatrace}
\end{figure}

\steven{TODO make the DFA parser figure fit}

The linear type of traces through $D$ is given in \cref{fig:dfatrace}. These
traces are either terminated in a state $s$ with the $\nil$ constructor. When
terminating a trace, we tag it with a Boolean that marks if the end state was
accepting or not. The $\cons$ constructor builds a trace out of state $s$ by linearly combining
a parse of some character $c$ with a
trace out of the state $D.\delta_{c}(s)$. The trace built with $\cons$ is
accepting if and only if the trace out of $D.\delta_{c}(s)$ is accepting.

Because DFAs are deterministic, we are able to prove that their type of traces
are unambiguous and define a parser for them. In particular we show that
$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}(s , b)}$ is isomorphic to
$\stringg$, which simultaneously constructs a parser and demonstrates that
traces through $D$ is unambiguous. This isomorphism intuitively says that we pay
unambiguously build traces through $D$ beginning at any state $s$, and further
that from any trace through $D$ we can uniquely recover the string that built
the trace.

The first half of the isomorphism,
$\parse_{D} : \ltonl {\left( \stringg \lto \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \right)}$,
is defined by induction on strings in \cref{fig:dfaparser}. If this string is empty, then $\parse_{D}$
defines a linear function that terminates a trace at the input state $s$. If the
string is nonempty, then $\parse_{D}$ walks forward from the input state $s$ by
the character at the head of the string.

\begin{figure}
\[
\inferrule
{
  \cdot \vdash mt := \withlamb {s} {\sigma~(D.\isAcc(b))~(\nil.\pi~s)}   : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool} {\Trace_{D}(s, b)}}
  \\
  ch : \charg , tr : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}} \vdash
  f := \withlamb {s} {\letin {\sigma~c~p_{c}} {ch} {\letin {\sigma~b~t} {tr(D.\delta_{c}(s))} {\sigma~b~(\cons(p_{c} , t))}}}
  : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}}
}
{w : \stringg \vdash \parse_{D} := \fold(mt, f)(w) : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}}}
\]
\caption{Parser for DFA traces}
\label{fig:dfaparser}
\end{figure}

The second half of the isomorphism,
$\print_{D} : \ltonl {\left( \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \lto \stringg \right)}$,
is given in \cref{fig:printdfa} by induction on traces. If the trace is defined
via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
is defined by $\cons$ then $\parse_{D}$ appends the character $c$ to the output
string and recurses.

\begin{figure}
\[
\inferrule
{
}
{tr : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}} \vdash asdf : \stringg}
\]
\caption{Parser for DFA traces}
\label{fig:printdfa}
\end{figure}

@@

\section{Denotational Semantics}
\label{sec:semantics-and-metatheory}

To justify our intuitive explanation of linear types as formal
grammars, we will now define a \emph{denotational semantics} that
makes this mathematically precise. This semantics is the basis for our
prototype implementation in Agda.

\subsection{What is a formal grammar?}

\begin{definition}
  A \emph{formal language} $L$ is a function from strings to propositions.
  A \emph{formal grammar} $A$ is a function from strings to small sets.
\end{definition}

\max{todo: need to decide if we need this section here or if it's in the intro already}.

@@

\begin{definition}

\end{definition}

To handle our universe, we assume a universe of small sets in our
metatheory that is closed under all of our type formers.
\begin{definition}[Grammar Semantics]
  We define the following interpretations by mutual recursion on the
  judgments of \theoryabbv:
  \begin{enumerate}
  \item For each non-linear context $\Gamma \isCtx$, we define a set $\sem \Gamma$.
  \item For each non-linear type $\Gamma \vdash X \isTy$, and element
    $\gamma \in \sem\Gamma$, we define a set $\sem X \gamma$.
  \item For each linear type $\Gamma A \isLinTy$ and element $\gamma
    \in \sem\Gamma$, we define a formal grammar $\sem{A}\gamma$. We
    similarly define a formal grammar $\sem\Delta\gamma$ for each
    linear context $\Gamma \Delta\isLinCtx$.
  \item For each non-linear term $\Gamma \vdash M : X$ and $\gamma \in \sem{\Gamma}$, we define an element $\sem{M}\gamma \in \sem{X}\gamma$.
  \item For each linear term $\Gamma; \Delta \vdash e : A$ and $\gamma \in \sem{\Gamma}$ we define a parse transformer from $\sem{\Delta}\gamma$ to $\sem{A}\gamma$.
  \end{enumerate}
  And we verify the following conditions:
  \begin{enumerate}
  \item If $\Gamma \vdash X \isSmall$, then $\sem X \gamma$ is a small set.
  \item If $\Gamma \vdash X \cong X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
  \item If $\Gamma \vdash A \cong A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
  \item If $\Gamma \vdash M \cong M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
  \item If $\Gamma;\Delta \vdash e \cong e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
  \end{enumerate}
\end{definition}

For reasons of brevity, we present only a selection of cases here. The full details are in the appendix.\max{todo: put the full details in the appendix}.

Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
depth the linear semantics.
\begin{enumerate}
  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
        $\sem X : \sem \Gamma \to \Set_{i}$
  \item A non-linear term $\Gamma \vdash e : X$ denotes a dependently typed function
        $\sem e : \prod_{\gamma : \sem \Gamma}\sem{X} \gamma$ \pedro{we have not defined what sections are}
  \item Linear contexts $\Gamma \vdash \Delta$ and types
        $\Gamma \vdash A : L_{i}$ both denote families of grammars
        $\sem \Gamma \to \Gr_{i}$
  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
        transformers
        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
\end{enumerate}
